FINE SCALE REGISTRATION OF SIDEWALKS AND OTHER RIBBON-LIKE FEATURES FROM AERIAL IMAGERY

### 2018-11-06 Chapters/1.Introduction/Introduction.tex
Corrections and updates to notation

\chapter{Introduction}

\section{Motivation}
Aerial imagery is an important resource for applications that include urban planning and map-making~\cite{5523977}. 
Ribbon-like features such as sidewalks, street networks, canals, or biking paths are especially important because they provide essential connections between sites to support transportation of people and materials. 
Any damage or obstacle that appears on these ribbon-like features may interfere with the navigability of the walking path and could even cause harm to people.
High-resolution aerial imagery has become a way to capture the appearance of these structures in order to look for changes such as new construction, degradation, damage, or obstructions. 
Geometrically, walking paths are thin features that can be described by a trajectory with varying thickness~\cite{10.1007/11744078_9}. 
The scope of this thesis is automatic refinement of ribbon-like features: given an approximate trajectory and range of potential thicknesses, we automate the process of estimating the thicknesses and offset of the trajectory. 
One example of a downstream application is to enable accurate walking instructions for people including those with disabilities that prevent them from moving around or seeing obstacles~\cite{ZOU2012227}.

\begin{figure}
    \centering
    \includegraphics[width=0.85\textwidth]{Figures/Goal.png}
    \caption[Objective Demonstration]{A demonstration  on input data and output result. With the input map and initial trajectory (left), we produce the precious boundaries(right) with our \ac{DP} approach. The trajectory mark in black and the boundaries mark in red. We output the edges information as a .geojson file which allow us to edit and change with the given input file.}
    \label{fig:goal}
\end{figure}

Public domain or crowd-sourced repositories of spatial data such as \ac{OSM}~\cite{OpenStreetMap} represent a very complete and frequently updated source of information which includes street vector data, and to a much lesser extent, the locations of foot-ways.
However, the data is acquired from a variety of sources that may not be registered with enough precision to accurately place thin features such as walking paths. For thin structures such as sidewalks that are visible in \ac{VHR} imagery, even small discrepancies between the digitized pathways and the imagery can be distracting when used to visualize an overlaid map.
As new \ac{VHR} aerial imagery becomes available thin features such as walking paths are discernible with higher precision, but ortho-rectification processes may not precisely align the imagery with previously digitized features.

@@

In addition, poorly registered walking paths could confound any algorithm that might hope to automatically locate features on the path including obstacles, damage, overgrowth, or stairways that could impact the walk-ability. 
Several authors\cite{femiani2009interval, femiani2007road} have worked on segmenting streets and sidewalks or other linear features directly from aerial imagery; however sidewalks are much thinner than streets and are often partially occluded. 
More importantly, these methods are mainly concerned with the connectivity of the extracted pathways and not with identifying the precise boundaries. 
Figure \ref{fig:goal} shows the input map with the initial trajectory (left) and result from our approach (right).
We marked initial trajectory in black and boundaries in red.
Figure \ref{fig:fw_ov} demonstrates the step by step detail diagram for our approach, from the input map (a), along with the initial trajectory(b), we can generate the ribbon-image with the given mid-curve(c). 
We generate the initial estimation for each pixels(d) with initial density along the ribbon-image, then apply our approach with \ac{DP} solution(e). 
Our approach, we name it \ac{DPST}, can be used to refine or recover the thickness of street networks or other networks of ribbon-like features. Several automatic approaches model segmentation as a \ac{CRF} in order to ensure that the resulting output configurations are plausible \cite{ActiveContou09, Rother2004-ou, Achanta:149300}.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{Figures/diagram.png}
    \caption[Framework Demonstration]{
    Our approach adopted to predict precise boundaries for ribbon-like features. Where black line indicated a sidewalk's geometric information, and red lines shown our result for the sidewalk boundaries. From (a) to (f), the 6 step approach introduce input map (a), initial trajectory (b), ribbon image generation (c), density Estimation for pixels(d), the result from our \ac{DP} solution(e), the result after reshape sidewalk(f).}
    \label{fig:fw_ov}
\end{figure}

\section{Challenges}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\textwidth]{Figures/Challenge.png}
    \caption[Challenges Demonstration]{A detail demonstrations of challenges that we are facing in our approach. From left to right, top to bottom, it shows that sidewalk are under the shadow, camouflage with adjacent materials and blocked by obstacles.}
    \label{fig:challenge_demo}
\end{figure}

@@

Under the circumstances, most feature segmenting tools will not able to locate accurate boundaries, their results may recognize the non-sidewalk pixels (FP) as the sidewalk or the other way around (FN).

\section{Contributions}

The aim of this work is to precisely align and determine the width of a coarsely registered ribbon-like feature to match its appearance in a \ac{VHR} image.
The input is a \ac{VHR} georeferenced and orthorectified image along with a set of linear features that are close to, but perhaps not precisely aligned with the center of a ribbon-like feature. 
In particular, we expect that the ribbon will be made of a material with a somewhat uniform appearance (e.g. asphalt, gravel, or concrete) but that its appearance will be affected by shadows and interrupted by occlusions from vegetation, pedestrians, or vehicles such as bicycles. 
Our aim is to determine a more precise mid-line and an estimate for the width of the ribbon-like feature, which can vary slowly along its trajectory. 
The solution proposed in this paper does not rely on prior knowledge of the material on or around the ribbon, but instead, it estimates material appearance as it refines the placement of an initial estimate for the ribbon's mid-line.

	Our contributions are as follows:
	\begin{itemize}
		\item A novel \ac{DP} approach is proposed for aligning ribbon-like features to orthoimagery.
		\subitem We do not require large training sets; a few parameters to control the smoothness of the resulting ribbon along with a coarse initial estimate of the trajectory are all that is needed.
		\subitem Unlike methods that aim to recover the trajectories only, in our approach a likely thickness is directly identified so that more plausible trajectories and boundaries can be identified.
		\item An application to the problem of registering sidewalks to aerial photographs is demonstrated, allowing \ac{GIS} data sets to be improved as high-resolution ortho imagery becomes available. 
	\end{itemize}

>>>>

\chapter{Introduction}\label{sec:introduction}

\section{Motivation}\label{sec:motivation}
Aerial imagery is an important resource for applications that include urban planning and map-making~\cite{5523977}.
Ribbon-like features such as sidewalks, street networks, canals, or biking paths are thin thin structures with a medial
curve and varying width. They are especially important because they provide essential connections between sites to
support transportation of people and materials. Any damage or obstacle that appears on these ribbon-like features may
interfere with the navigability of the walking path and could even cause harm to people. High-resolution aerial imagery
has become a way to capture the appearance of these structures in order to look for changes such as new construction,
degradation, damage, or obstructions. Geometrically, walking paths are thin features that can be described by a
trajectory with varying thickness~\cite{10.1007/11744078_9}. The scope of this thesis is automatic refinement of
ribbon-like features: given an approximate trajectory and range of potential thicknesses, we automate the process of
estimating the thicknesses and offset of the trajectory. One example of a downstream application is to enable accurate
walking instructions for people including those with disabilities detect and avoid obstacles such as stairways~\cite{ZOU2012227}.

\begin{figure} 
\centering 
\includegraphics[width=0.85\textwidth]{Figures/Goal.png} 
\caption[Objective Demonstration]{
A demonstration  on input data and output result. 
With the input map and initial trajectory (left), 
we estimate detailed boundaries(right) with our \ac{DP} approach. 
The trajectory is marked in black and the boundaries are marked in red. 
We output the edges information as a .geojson file.}
\label{fig:goal} 
\end{figure}

Public domain or crowd-sourced repositories of spatial data such as \ac{OSM}~\cite{OpenStreetMap} represent a very
complete and frequently updated source of information which includes street vector data, and to a much lesser extent,
the locations of walking paths (called foot-ways by \ac{OSM}). However, the data is acquired from a variety of sources that may not be registered with
enough precision to accurately place thin features such as walking paths. For thin structures such as sidewalks that are
visible in \ac{VHR} imagery, even small discrepancies between the digitized pathways and the imagery can be distracting
when used to visualize an overlaid map. As new \ac{VHR} aerial imagery becomes available thin features such as walking
paths are discernible with higher precision, but ortho-rectification processes may not precisely align the imagery with
previously digitized features.

@@

In addition, poorly registered walking paths could confound any algorithm that might hope to automatically locate
features on the path including obstacles, damage, overgrowth, or stairways that could impact the walkability. Several
authors\cite{femiani2009interval, femiani2007road} have worked on segmenting streets and sidewalks or other linear
features directly from aerial imagery; however sidewalks are much thinner than streets and are often partially occluded.
More importantly, these methods are mainly concerned with the connectivity of the extracted pathways and not with
identifying the precise boundaries.

The problem we solve is illustrated in Figure \ref{fig:goal}.  We are given an input orthorectified image with an
initial trajectory (left) and we aim to find a ribbon (right). The initial trajectory is drawn in black, and boundaries 
of the ribbon are drawn in in red. 
Figure \ref{fig:fw_ov} demonstrates a step by step  diagram of our approach, 
from the input map (a), along with the initial
trajectory (b), we generate the ribbon-image with the given medial-curve (c).
We generate an initial estimation for the likelihood of each
pixel (d) with along the ribbon-image, then apply our \ac{DP} solution (e)
 to find a \ac{MAP} estimate of the ribbon.
Our approach, which we call \ac{DPRST}, can be used to refine the trajectories or recover the thickness of 
walking paths or other ribbon-like features. 
%
%Several automatic approaches model segmentation as a \ac{CRF} in order to ensure that the resulting output
%configurations are plausible \cite{ActiveContou09, Rother2004-ou, Achanta:149300}.

\begin{figure}[H] 
\centering 
\includegraphics[width=\textwidth]{Figures/diagram.png} 
\caption[Framework Demonstration]{
Our approach adopted to predict precise boundaries for ribbon-like features. 
The black line strip indicates a sidewalk's geometric information, 
and red lines show the identified ribbon boundaries. 
From top-left to bottom-right we show:
(a)~the input map, 
(b)~initial trajectory, 
(c)~the warped ribbon image, 
(d)~density estimation for the likelihoods of pixels, 
(e)~the result from our \ac{DP} solution, 
(f)~the result after reshaping sidewalk.} 
\label{fig:fw_ov} 
\end{figure}

\section{Challenges}\label{sec:challenges}

\begin{figure}[H] 
\centering 
\includegraphics[width=0.75\textwidth]{Figures/Challenge.png} 
\caption[Challenges Demonstration]{
A demonstration of some challenges faced by in our approach. From left to right, top to
bottom, it shows 
a sidewalk whose appearance is interrupted by shadow,
camouflage with adjacent materials,
and blocked by obstacles.}
\label{fig:challenge_demo}
\end{figure}

@@

Under these circumstances, most feature segmenting tools will not able to locate accurate boundaries, their results may
recognize the non-sidewalk pixels as the sidewalk or the other way around.

\section{Contributions}\label{sec:contributions}

The aim of this work is to precisely align and determine the width of a coarsely registered ribbon-like feature to match
its appearance in a \ac{VHR} image. The input is a \ac{VHR} georeferenced and orthorectified image along with a set of
linear features that are close to, but perhaps not precisely aligned with the center of a ribbon-like feature. In
particular, we expect that the ribbon will be made of a material with a somewhat uniform appearance (e.g. asphalt,
gravel, or concrete) but that its appearance will be affected by shadows and interrupted by occlusions from vegetation,
pedestrians, or vehicles such as bicycles. Our aim is to determine a more precise mid-line and an estimate for the width
of the ribbon-like feature, which can vary slowly along its trajectory. The solution proposed in this paper does not
rely on prior knowledge of the material on or around the ribbon, but instead, it estimates material appearance based on
on the initial trajectory. We
refine the placement of an initial estimate for the ribbon's mid-line with a shape-based prior that favors a result with little 
change in direction (bending) of the medial trajectory, and that favors small changes in thickness.

Our contributions are as follows: 
\begin{itemize} 
\item We introduce a \textit{Ribbon Prior} for segmenting linear feature that have slowly varying
      width and smooth (low curvature) medial trajectories. 
\item A novel \ac{DP} approach is proposed for aligning ribbon-like features to orthoimagery. 
    We do not require large training sets; a few parameters to control the smoothness of
	the resulting ribbon along with a coarse initial estimate of the trajectory are all that is needed. 
	Unlike
	methods that aim to recover the trajectories only, in our approach a likely thickness is directly identified so that
	more plausible trajectories and boundaries can be identified. 
\item An application to the problem of registering
	sidewalks to aerial photographs is demonstrated, allowing \ac{GIS} data sets to be improved as high-resolution ortho
	imagery becomes available. 
\end{itemize}

### 2018-11-06 Chapters/2.Theory/Background.tex
Corrections and updates to notation

\chapter{Background}

% To have a better understanding of our approach, it's important to understand the basic concept and assumption that we used to develop the approach.

\section{\ac{VHR} Aerial Imagery}

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{Figures/arz_7_res_compare.png}
    \caption[Resolution Comparison]{Comparison between different resolutions. From left to right, where (a) has half meter scales, (b) has 5 meter scales and (c) has 10 meter scales. With the increase of meter scale, it harder to capture the boundary of the edges detail of sidewalks.}
    \label{fig:vhr_compare}
\end{figure}
% \todo{change dpi to 5 meter scale. make it worse.}

\ac{VHR} stands for Very High Resolution, it has advantageous for buildings extraction, road detection or image segmentation~\cite{FREIRE20141, 10.1007/BFb0015525, 10.1007/978-3-642-15567-3_16}. 
These aerial imageries contain more information and detail due to the higher resolution, which give us the ability to develop more complex analysis on the imagery with same size.
\figref{fig:vhr_compare} demonstrates the difference between different resolution within the same area. From (a) to (c), the resolution changes from half meter scale to 10 meter scale. 
With higher resolution, it's easier to find the precious boundaries of an object or identify it. 
For building extraction, higher resolution provides more information when dealing with complex houses or extract complex suburban roofs~\cite{10.1007/BFb0015525}. 
For thin ribbon-features segmentation such as sidewalks in our approach, to visualize an overlaid map without high resolution, the imagery can be distracting, and the output may effect by tiny discrepancies between digitized pathways~\cite{10.1007/978-3-642-15567-3_16}.

\section{\ac{OSM}}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\textwidth]{Figures/oxford_path_data_osm.png}
    \caption[\ac{OSM} Walking Path]{A demonstration of actual walking path data from \ac{OSM}. The figure shows in the Oxford area in Ohio, US. Row 2 specifies the zoom in version of error that we need to facing. The sidewalk misses in the path data (left), the path data marks offset compare to the actual sidewalk(right).}
    \cite{OpenStreetMap}
    \label{fig:osm_oxford_path}
\end{figure}

\ac{OSM} stands for Open Street Map, which was founded by Steve Coast in 2004~\cite{lasPiñas, OpenStreetMap}. 
\ac{OSM} is a collaborative project that allows users to generate a free editable map of the world~\cite{4653466}.
Rather than the map itself, \ac{OSM} generates the data with the map information that across most of the world as primary output. 
It also allows users to develop the map data with portable satellite navigation devices. 
With all these convenience features, \ac{OSM} is helpful for the map analysis and features segmentation~\cite{10.1007/11744078_9}. 
\figref{fig:osm_oxford_path} shows that \ac{OSM} supports features like footpaths, roads, buildings, and others, with the ability to extract the \ac{GIS} data for any individual feature. 
It may also leads to the issue of dealing with missing or invalid data when we rely on the data set for auto-detection, or mismatch data with imagery when structures changes along with time.

\section{\ac{CRF}}
\ac{CRF} stands for Conditional Random Field, which is a class of statistical modeling method that often applied in pattern recognition and machine learning~\cite{MAL-013}. 
As a type of discriminative undirected probabilistic graphical model, \ac{CRF} is commonly used for segmenting, labeling or parsing of sequential data, such as in computer vision, structure segmentation and object recognition~\cite{RuizSarmiento2015UPGMppA, 1315232, lafferty2001conditional}. 
For road segmentation, several approaches are able to locate the precise boundaries by treating it as a \ac{CRF} problem~\cite{ActiveContou09, Achanta:149300}.

\section{\ac{GMM}}

\ac{GMM} stands for Gaussian Mixture Models~\cite{sridharan2014gaussian}.
It is a probabilistic model for representing normally distributed subpopulations within an overall population.
\ac{GMM} often can be performed when the data points are mixtures of a Gaussian distribution. 
It can be used in common computer vision tasks such as background and foreground subtraction~\cite{1333992}. 
For road segmentation, we can treat the problem as subtract foreground (road feature) from the complex background (non-road feature). 
\figref{fig:gmm_sample_1} demonstrate the comparison between foreground and background \ac{GMM} on sample walking path show in gray-scale. 
For foreground subtraction, which shown in row 1 and row 3, we treat sidewalk texture as foreground and apply the \ac{GMM} algorithm to subtract it from the background (non-sidewalk texture). 
Background subtraction are shown in row 2 and row 4, we did the opposite and treat non-sidewalk part as foreground.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\textwidth]{Figures/GMM_needed.png}
    \caption[Density Background Subtraction]{A demonstration of apply \ac{GMM} algorithm on sample inputs, show with gray scale. From top to bottom, row 1 shows foreground density on sample path, row 2 show background density on the same sample path as row 1. Same with row 3 and 4, with different sample path. For foreground density, lighter color represent more likely to be sidewalk and darker color represent more likely to be non-sidewalk. Background density shows as the opposite.}
    \label{fig:gmm_sample_1}
\end{figure}

\section{\ac{DP}}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{Figures/longest_common_sequence.png}
    \caption[Demonstration on Longest Common Sequence]{Dynamic Programming demonstration with longest common sequence. Similar to our approach, to find the longest common sequence, it first needs to fill the table with \ac{DP} and then backtrack to find the longest path. It used in intersection detection with GPS trace~\cite{Xie2017DetectingRI}.
    }
    \label{fig:lcs_dp}
\end{figure}

\ac{DP} stands for Dynamic Programming. It is both a mathematical optimization method and a computer programming method \cite{bertsekas2005dynamic}. 
In 1950s \ac{DP} was invented by Richard Bellman\cite{bellman2013dynamic}. 
The main concept of \ac{DP} is to simplify a complicated problem by breaking it down into simpler sub-problems in a recursive manner \cite{howard1966dynamic}. 
In computer science, if a problem can break into serial sub-problems and each sub-problem can be recursively solved by finding its optimal solution, then we know that \ac{DP} are applicable.

\figref{fig:lcs_dp} gives a example of a famous dynamic programming problem, Longest Common Sequence. To achieve the result of finding the longest common sequence, it first needs to fill the table with \ac{DP} and then backtrack to find the longest path. One method gives approach with using longest common sequence for intersection detection from GPS traces~\cite{Xie2017DetectingRI}.

Dynamic Programming has a variety of applications since it can significantly reduce the run-time \cite{bertsekas1995neuro}. 
We use Fibonacci Sequence to demonstrate the \ac{DP} process\cite{horadam1961generalized}. 
Figure \ref{fig:fbs}, row 1 shows that to calculate the $n^{th}$ Fibonacci number, it require a certain exponential rate of calculation to finish the process.
The time complexity can be reduced since it is re-calculating the previous Fibonacci number repeatedly. 
By applying \ac{DP} algorithm, we can save the previous calculated Fibonacci number and reused it when needed.
Figure \ref{fig:fbs}, row 2 indicate the same process after applying dynamic programming, we marked unnecessary steps in red X. 
By trade in linear space to save previous calculation, we can reduce the wrong time from exponential to linear.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{Figures/treewithdp.png}
    \caption[Demonstration on Fibonacci Sequence]{Representing the calculation of Fibonacci Sequence ($7^{th}$) with binary tree. Row 1 shows without applying dynamic programming, it need's certain exponential rate to calculate $n^{th}$ Fibonacci number. With the increase of $n$, the number of calculation increase significantly. Row 2 shows with dynamic programming, after saving previous calculation, it's only require linear time steps to calculate $n^{th}$ Fibonacci number. We marked unnecessary steps in red X to show it's can be skipped. }
    % \cite{stack_overflow}
    \label{fig:fbs}
\end{figure}

>>>>

\chapter{Theoretical Background}

In order to have a better understanding of our approach, it's important to understand concepts 
related to \ac{VHR} orthorectified imagery (\S\ref{sec:vhr}),
\ac{GIS} data and in particular the the \ac{OSM}  project (\S\ref{sec:osm}), 
\acp{CRF} and their use in \ac{MAP} image segmentation (\S\ref{sec:crf}),
density estimation especially with \acp{GMM} (\S\ref{sec:gmm}), and 
the algorithm design technique of \ac{DP} (\S\ref{sec:dp}).

\section{\ac{VHR} Aerial Imagery}\label{sec:vhr}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{Figures/arz_7_res_compare.png}
\caption[Resolution Comparison]{
Comparison between different resolutions. 
From left to right, where
(a) has 10cm between samples, 
(b) has 0.5 meter spacing and 
(c) has 1.0 meter spacing. 
With the decrease in resolution,
it becomes harder to capture the boundary detail of the edges  of sidewalks.}
\label{fig:vhr_compare}
\end{figure}
% \todo{change dpi to 5 meter scale. make it worse.}

\ac{GSD} is the distance on the ground between the centers of pixels in an aerial image; it is
inversely related to the resolution of the image. \ac{VHR} imagery captures photographic details
with a vary small \ac{GSD} -- often less than one meter. Much of the data we target is at
resolutions well below one meter between samples; for example \ref{fig:vhr_compare}(a) has \ac{GSD}
of 10cm. The ability to resolve details at a small \ac{GSD} is advantageous for building
extraction, road detection or image segmentation~\cite{FREIRE20141, 10.1007/BFb0015525,
10.1007/978-3-642-15567-3_16}, although with additional detail one often encounters noise
or clutter that may interfere with object segmentation. \ac{VHR} aerial
images contain more information and detail due to the higher resolution, which gives us the ability
to develop more complex analysis on the imagery. \figref{fig:vhr_compare}
demonstrates the difference between different resolutions within the same area. From (a) to (c), the
resolution changes from $\frac{1}{10}$ meter \ac{GSD} to 1.0 meter \ac{GSD}. With higher resolution,
it is easier to find the precise boundaries of an object or even identify thin objects at all. For
building extraction, higher resolution provides more information when dealing with complex roofs
~\cite{10.1007/BFb0015525}. For thin ribbon-features segmentation such
as sidewalks in our approach, the output may be effected by tiny discrepancies between digitized
pathways~\cite{10.1007/978-3-642-15567-3_16}. Aerial images used to make digital maps have often
been subjected to  a process called \textit{ortho rectification}, in which a digital elevation model
has been used to predict the parallax offset between pixels in the image and their position on a
map. Perspective causes objects at higher elevation to appear to shift away from the center of an
image. Ortho-rectification predicts the distortion and corrects it to make the image appear as if it
was taken by an orthographic camera. However, most elevation models are very coarse compared to the
thin features that can be seen in \ac{VHR} images, and small errors in the ortho-rectification
process may cause thin or small features to appear to shift. It is also common for images to be
subjected to resolution ``enhancing'' operations such as pan-sharpening that attempt to artificially
increase the apparent resolution of the image. Prior image processing and rectification operations are 
useful but they can also challenge attempts to segment objects based only on object material or texture.

\section{\ac{OSM}}\label{sec:osm}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\textwidth]{Figures/oxford_path_data_osm.png}
    \caption[\ac{OSM} Walking Path]{A demonstration of actual walking path data from \ac{OSM} \cite{OpenStreetMap}. 
    The figure shows the campus of Miami University in Oxford, Ohio in USA. 
    Row 2 shows a zoomed version of the imagery.
    The sidewalk misses in the path data (left), the path is offset compared to the actual sidewalk(right).}

    \label{fig:osm_oxford_path}
\end{figure}

\ac{OSM} stands for Open Street Map, which was founded by Steve Coast in 2004~\cite{lasPiñas, OpenStreetMap}. 
\ac{OSM} is a collaborative project that allows users to generate a free editable map of the world~\cite{4653466}.
%Rather than the map itself, \ac{OSM} generates the data with the map information that across most of the world as primary output. 
It also allows users to develop the map data with portable 
satellite navigation devices, for example by recording GPS trails. 
With all these convenient features, \ac{OSM} is a helpful tool for  map analysis
 and feature segmentation~\cite{10.1007/11744078_9}. 
\figref{fig:osm_oxford_path} shows that \ac{OSM} supports features such as footpaths, roads, buildings, 
and others. There are public APIs that provide users with the ability to extract the \ac{GIS} data for 
any individual feature. 
Although crowd-sourced map data is a powerful resource, it may also lead to the issue of
missing, inaccurate or invalid data when we rely on the data set for sidewalk localization. 
In addition any GIS data may no longer match imagery as structures change over time.

\section{\ac{CRF}} \label{sec:crf}
\acfp{CRF} are is a class of statistical graphical modeling methods that often applied in pattern recognition and machine learning~\cite{MAL-013}. 
As a type of discriminative undirected probabilistic graphical model, \acp{CRF} are commonly used for segmenting, labeling or parsing of sequential data, such as in computer vision, structure segmentation and object recognition~\cite{RuizSarmiento2015UPGMppA, 1315232, lafferty2001conditional}. 
For road segmentation, several approaches are able to locate the precise 
boundaries by treating it as a \ac{CRF} problem~\cite{ActiveContou09, Achanta:149300}.

\section{\ac{GMM}}\label{sec:gmm}

\acfp{GMM} represent probability densities as a mixture of (e.g. multivariate) normal distributions 
~\cite{sridharan2014gaussian}. A \ac{GMM} is a probabilistic model for representing normally
distributed subpopulations within an overall population. A \ac{GMM} can often  be estimated using
the \ac{EM-GMM} algorithm based on data points which are sampled from a mixture of Gaussian distributions. It can
be used in common computer vision tasks such as background and foreground
subtraction~\cite{1333992}. For road segmentation, we can treat the problem as separating foreground
(colors within a ribbon) from the complex background (non-ribbon colors). \figref{fig:gmm_sample_1}
demonstrates the comparison between foreground and background \acp{GMM} on a sample walking path,
with the density estimate for each pixel shown in gray-scale. For foreground density estimation,
which we show in row 1 and row 3, we treat sidewalk colors as foreground and apply the \ac{EM-GMM}
algorithm to estimate the probability density $\Pr(\Color{}|R)$ where $\Color{}$ is the event that a
color is observed in the image and $R$ is the event that we are part of a sidewalk (ribbon).
Background likelihoods are shown in row 2 and row 4, we did the opposite and estimated
$\Pr(\Color{}|\neg R)$. More details including the number of Gaussian distributions in each mixture
we use are described in section \S\ref{sec:density-estimation}, and the result of using Bayes theorem to estimate the posterior probability 
$\Pr(R|\lambda)$ that a pixel is part of a ribbon are shown in \figurename{}\ref{fig:GMM_Sample_2}

\begin{figure}[H]
\centering
\includegraphics[width=0.85\textwidth]{Figures/GMM_needed.png}
\caption[Density Background Subtraction]{
A demonstration of the \ac{EM-GMM} algorithm on sample inputs, probability densities are shown in gray scale. 
From top to bottom, row 1 shows foreground density on a sample path, row 2 show background density on 
the same sample path as row 1. 
Row 3 and 4 show the same with a different sample path. 
A lighter color represents higher probability density.}
\label{fig:gmm_sample_1}
\end{figure}

\section{\ac{DP}}\label{sec:dp}

\acf{DP} is both a mathematical optimization method and a computer programming method \cite{bertsekas2005dynamic}. 
In 1950s \ac{DP} was invented by Richard Bellman\cite{bellman2013dynamic}. 
The main concept of \ac{DP} is to simplify a complicated problem by breaking it down into smaller overlapping sub-problems
 in a recursive manner \cite{howard1966dynamic}. 
In computer science, if a problem can divided into several overlapping sub-problems, 
each sub-problem can be recursively solved by finding its optimal solution, and if some subproblems are repeated, then 
we know that \ac{DP} is applicable. \ac{DP} is characterized by caching or \textit{memoizing}
 the solutions to repeated subproblems in a table, and the using the stored values rather than solving
 the same problem multiple times.

Dynamic Programming has a variety of applications since it can significantly reduce the 
running-time of some algorithms \cite{bertsekas1995neuro}. 
We use Fibonacci Sequence 
$$F(n) = \begin{cases}
1 & \text{if $n=1$ or $n=2$}\\
F(n-1)+F(n-2) & \text{if $n>2$}
\end{cases}$$
to demonstrate the \ac{DP} process\cite{horadam1961generalized}. 
Figure \ref{fig:fbs}, row 1 shows that to calculate the $n^{th}$ Fibonacci number, one must 
evaluate exponentially growing number of subproblems to finish the process.
The time complexity can be reduced since we are re-calculating certain Fibonacci numbers repeatedly. 
By applying a \ac{DP} approach, we can save the previous calculated Fibonacci number and reuse it when needed.
Figure \ref{fig:fbs}, row 2, shows the same process after applying dynamic programming. Unnecessary steps are marked with a red `X'. 
By trading a linear amount of space (for caching results) to save previous calculation, we can reduce the running time from exponential to linear.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{Figures/treewithdp.png}
\caption[Demonstration on Fibonacci Sequence]{
In image from \cite{stack_overflow} representing the calculation of the $7^{th}$ Fibonacci number with a recursion tree. 
Row 1 shows without applying dynamic programming, one must solve an exponential number of subproblems to calculate $n^{th}$ Fibonacci number. 
With the increase of $n$, the number of calculations increase significantly.
Row 2 shows that with dynamic programming, after saving previous calculations, 
one requires a linear number of steps (left edge of the tree) 
to calculate $n^{th}$ Fibonacci number.
We marked unnecessary steps in red ``X'' to show it can be skipped. }

\label{fig:fbs}
\end{figure}

\ac{DP} often involves first filling in a table with the \textit{optimal value} of sub-problems, and
in order to recover the sequence of decisions that led to the optimal value a process of \textit{backtracking} is used. 
\figref{fig:lcs_dp} gives a example of a famous dynamic programming problem, finding the length of the Longest Common Sequence. 
Given two strings $t_1=\langle t_{1,1}, t_{1,2}, \dots, t_{1,m}\rangle$
and $t_2=\langle t_{2,1}, t_{2,2}, \dots, t_{2,n}\rangle$, we aim to identify 
the \textit{length} of the longest subsequence that is common to both $t_1$ and $t_2$.
In this case a subproblems are prefixes of the strings $t_1$ and $t_2$.
Once we know the lengths of the longest subsequences for every subproblem, we can use 
backtracking to generate the longest subsequence itself. 
The longest common sequence is useful for geospatial processing, for example it has been use for intersection detection from GPS
traces~\cite{Xie2017DetectingRI}.

The length of the $\mathtt{LCS}$ for prefixes of length $i$ and $j$ is determined recursively as
$$
\mathtt{LCS}(i, j) = \begin{cases}
\mathtt{LCS}(i - 1, j - 1) + 1 & \text{if $t_{1, i} = t_{2,j}$}\\
\max\left[
\mathtt{LCS}(i-1, j),
\mathtt{LCS}(i, j-1)
\right] 
& \text{else}\\
\end{cases}
$$
and a subproblem is solved for each pair of sizes $i,j$ in the table of \figref{fig:lcs_dp}.
As each subproblem is solved, one of three choices was made in the recurrence (depicted by arrows in the table).

By following the arrows in \textit{reverse} starting at $m,n$, one is able to determine which
 symbols were common to both strings in the longest subsequence.

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figures/longest_common_sequence.png}
\caption[Demonstration on Longest Common Sequence]{
Dynamic Programming demonstration with longest common sequence from \cite{Xie2017DetectingRI}. 
Similar to our approach, to find the longest common sequence, one first fills the table using \ac{DP} and
 then backtracks to find the optimal path. 
}
\label{fig:lcs_dp}
\end{figure}
